---
title: "Optimization"
author: "Math 5530-4530-Spring 2024"
date: "2024-04-11"
output:
  html_document: default
  word_document: default
---




### Newton -Rapson method for finding maximum or minimum value of a function


```{r}
## max
newton=function(ftn,x0,tol=0.00001,iter.max=100){
  x=x0;
  fx=ftn(x);
  iter=0;
  while(abs(fx[2])>tol & (iter<iter.max)){
    x=x-fx[2]/fx[3]
    fx=ftn(x)
    iter=iter+1
    cat("number of iterations is",iter,"\n")
  }
  if(iter>=iter.max){
    cat("Algorithm failed to converge: try again \n")
    return(NULL)
  }else{
    cat("Algorithm converges \n")
    val=ftn(x)
    if(val[3]>0){
      cat("Second derivative is positive. So, the local minimum value is: ",x,"\n")
      return(x)
    }else{
      cat("Second derivative is negative. So, the local maximum value is: ",x,"\n")
      return(x)
    }
  }
}


```





### lets find the maximum value of $f(x)= -2x^2 +6x +5.$

#### Lets plot graph of the function and its derivative


```{r}
x=seq(-5,5,0.001)
y1=-2*x^2+6*x +5
y2=-4*x +6  ## Derivative
plot(x,y1,type="l",lty=5,bty="n",col="red",xaxt="n",yaxt="n",xlab="",ylab="")
xlabel=seq(-5,5,2)
xlabel=xlabel[xlabel!=0]
ylabel=seq(-100,50,10)
axis(1,xlabel,pos=0,lwd=1.5,col="dodgerblue")
axis(2,ylabel,pos=0,lwd=1.8,col="blue")
lines(x,y2,col="purple",lwd=2,lty=4)

```



```{r}
x=seq(-5,5,0.001)
y1=-2*x^2+6*x +5
y2=-4*x +6  ## Derivative
par(mfrow=c(1,2))
plot(x,y1,type="l",lty=5,bty="n",col="red",xaxt="n",yaxt="n",xlab="",ylab="",main="Graph of the function")

xlabel=seq(-5,5,2)
xlabel=xlabel[xlabel!=0]
ylabel=seq(-100,50,10)
axis(1,xlabel,pos=0,lwd=1.5,col="dodgerblue")
axis(2,ylabel,pos=0,lwd=1.8,col="blue")

plot(x,y2,type="l",lwd=1.5,lty=6,bty="n",col="purple",xaxt="n",yaxt="n",xlab="",ylab="",main="Graph of the derivative function")

xlabel=seq(-5,5,2)
xlabel=xlabel[xlabel!=0]
ylabel=seq(-100,50,10)
axis(1,xlabel,pos=0,lwd=1.5,col="dodgerblue")
axis(2,ylabel,pos=0,lwd=1.8,col="blue")

```


### The problem of finding maximum of a function redices to finding roots of its derivative.





```{r}
my_quad1=function(x){
  fx=-2*x^2+6*x+5
  dfx=-4*x+6
  dfx2=-4
  return(c(fx,dfx,dfx2))
}

```


### Apply Newton_Rapson Method

```{r}

newton(my_quad1,1)

newton(my_quad1,5)


```



### Example 2 Lets find the maximum and minimum of the function $2x^3-15x^2+36x+40$

#### lets plot the function and its derivative first

```{r}
x=seq(-1,4,0.001)
y1=2*x^3-15*x^2+36*x+40
y2=6*x^2 -30*x+36  ## Derivative
plot(x,y1,type="l",lwd=1.5,lty=5,bty="n",col="red",xaxt="n",yaxt="n",xlab="",ylab="")
xlabel=seq(-1,4,1)
xlabel=xlabel[xlabel!=0]
ylabel=seq(-100,100,10)
ylabel=ylabel[ylabel!=0]
axis(1,xlabel,pos=0,lwd=1.5,col="dodgerblue")
axis(2,ylabel,pos=0,lwd=1.8,col="blue")
lines(x,y2,col="purple",lwd=2,lty=4)
legend(2.5,40,legend = c("function","derivative"),col=c("red","purple"),lty=c(5,4),lwd=c(1.5,2),cex=0.8)

```


### Apply the Newton_Rapson method

```{r}
my_quad1=function(x){
  fx=2*x^3-15*x^2+36*x+40
  dfx=6*x^2-30*x+36
  dfx1=12*x-30
  return(c(fx,dfx,dfx1))
}


```


```{r}
newton(my_quad1,0)

```


```{r}
newton(my_quad1,5)

```

## Golden Section Method for optimization (maximum)

```{r}
golden_section=function(ftn, xl,xm,xr,tol=0.0000001){
  ## Check for inputs
  f.l=ftn(xl);
  f.m=ftn(xm);
  f.r=ftn(xr);
  
  
  if(f.l>=f.m){
    cat("error: f.l>=f.m \n")
    return(NULL)
  }
  
  if(f.r>=f.m){
    cat("error:f.r>=f.m \n")
    return(NULL)
  }
  # golden ratio plus one
  gr1 <- 1 + (1 + sqrt(5))/2;
  
  while(xr-xl>tol){
    if((xr-xm)>(xm-xl)){
      y=xm+(xr-xm)/gr1
      f.y=ftn(y)
      if(f.y>=f.m){
        xl=xm
        xm=y
        f.l=f.m
        f.m=f.y
      }else{
        xr=y
        xm=y
        f.r=f.y
      }
    }else{
      y=xm-(xm-xl)/gr1
      f.y=ftn(y)
      if(f.y>=f.m){
        xr=xm
        f.r=f.m
        xm=y
        f.m=f.y
      }else{
        xl=y
        f.l=f.y
      }
    }
  }
  cat("The function has maximum value at",xm,"\n")
  return(xm)
}


```


#### Example 1: find the maximum of $-2x^2+6x+5$

```{r}
my_quad=function(x){
  fx=-2*x^2+6*x+5
  return(fx)
}

```


```{r}
golden_section(my_quad,0,1,3)

```

###  Use the golden-section search algorithm to find all of the local maxima of the function 
$f(x)= 0$ if $x=0$

$f(x)= |x|\;\;log(\frac{|x|}{2})\;\;e^{-|x|}$ , otherwise


#### Lets plot the function first


```{r}
my_func3=function(x){
  y=vector()
  for ( i in 1:length(x))
  if (x[i] ==0){
    y[i]=0
  }else{
    y[i]= abs(x[i])*log(abs(x[i])/2)*exp(-abs(x[i]))
  }
  return(y)
}
my_func3(0.5)
```





```{r}
x=seq(-10,10,0.001)
y=my_func3(x)
plot(x,y,type="l",lty=5,bty="n",col="red",xaxt="n",yaxt="n",xlab="",ylab="",main="Graph of the function")

xlabel=seq(-10,10,1)
xlabel=xlabel[xlabel!=0]
ylabel=seq(-1,0.2,0.05)
ylabel=ylabel[ylabel!=0]
axis(1,xlabel,pos=0,lwd=1.5,col="dodgerblue")
axis(2,ylabel,pos=0,lwd=1.8,col="blue")


```



```{r}
golden_section(my_func3,2.1,2.3,5)
my_func3(3.170464)
```






$$\\[2 in]$$

















