---
title: "QUIZ 2"
author: "Nii Adjetey Adjei-Annan"
date: "2024-02-07"
output: html_document
---
```{r}
library(ggplot2)
```
```{r}
titanic= read.csv("titanic.csv")
titanic= na.omit(titanic)
titanic
```


###a. Calculate the frequency and relative frequency of survival for Titanic passengers
```{r}

table(titanic$Survived)
prop.table(table(titanic$Survived))
```
###b. Calculate the frequency and relative frequency of passengers in each ticket class.
```{r}
table(titanic$Pclass)
prop.table(table(titanic$Pclass))
```

#####c. Create a bar chart to display the number of passengers who survived and did not survive the Titanic sinking.
```{r}
ggplot(titanic, aes(x=Survived))+ geom_bar( col="brown", fill="yellow") + labs(title= "Survival Count Of Titanic Passenger", x = "Survived", y= "Passengers Onboard")
```

####d.Create a bar chart to display the percent of passengers who travelled under each ticket class.

```{r}
class_percent <- prop.table(table(titanic$Pclass)) * 100
class_percent_df <- data.frame(Pclass = as.factor(names(class_percent)), Percentage = as.numeric(class_percent))
ggplot(class_percent_df, aes(x = Pclass, y = Percentage)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Percentage of Passengers by Ticket Class",
       x = "Ticket class",
       y = "Percentage of Passengers") +
  scale_x_discrete(labels = c("1st Class", "2nd Class", "3rd Class")) +
  ylim(0, 100) +
 theme_minimal()
```

#####e.Create a contingency table to show the joint distribution of survival and passenger ticket class.

```{r}
table(titanic$Survived, titanic$Pclass)
```

###f. Draw a side-by-side bar chart to show the number of passengers in each ticket class and who survived and did not survive. 

```{r}
titanic$Pclass = factor(titanic$Pclass)
titanic$Survived =factor(titanic$Survived)
ggplot(titanic, aes(x = Pclass, fill = Survived)) +
  geom_bar(position = "dodge") +
  labs(title = "Passenger survival by Ticket Class",
       x = "Ticket class",
       y = "Number of passengers",
       fill = "Survived") +
  scale_fill_manual(values = c("yellow", "green"),
                    labels = c("Not_survived", "Survived")) +
 theme_minimal()
```

####Summarize your findings about survival on the Titanic and how a passenger’s ticket class might have affected his or her survival
```{r}
library(tidyverse)


```

####g.Summarize your findings about survival on the Titanic and how a passenger’s ticket class might have affected his or her survival
```{r}


survival_rate = titanic %>%
  group_by(Pclass) %>%
  summarize(survival_rate = mean(Survived == 1))

survival_rate

###From the data, passengers in 1st class has about 63% survival rate whereas passengers in the 3r class had the least surviving rate, which is about 25%. About half of passengers in the 2nd class had same surviving rate compared to the other half. 

```


